---
title: "01 - Forecasting live births"
format: html
editor: visual
toc: true
---

This script aims to describe the process of applying the method for forecasting the number of live births (LB) in each health macro-region in Brazil. To do this, we will use data from the Live Births System (SINASC, *Sistema de Nascidos Vivos*) provided by the SUS Information Technology Department (DATASUS) for each health macro-region.

# Preparing R environment

Loading libraries needed to forecast LB.

```{r}
#| warning: FALSE

library(tidyverse)
library(modeltime)
library(RODBC)
library(xgboost)
library(tidymodels)
library(timetk)
library(patchwork)

```

# Data extraction

We loaded the DATASUS data into our own infrastructure in order to access it using SQL, given the volume of data and the need to increase storage and processing efficiency.

The script below performs the data access.

```{r}
#| warning: false

dremio_host <- Sys.getenv("endereco")
dremio_port <- Sys.getenv("port")
dremio_uid <- Sys.getenv("uid")
dremio_pwd <- Sys.getenv("datalake")

channel <- 
        odbcDriverConnect(
                sprintf("DRIVER=Dremio Connector;
                        HOST=%s;PORT=%s;UID=%s;PWD=%s;
                        AUTHENTICATIONTYPE=Basic Authentication;
                        CONNECTIONTYPE=Direct", 
                        dremio_host, dremio_port, 
                        dremio_uid, dremio_pwd))

query <- 'SELECT * 
            FROM "Analytics Layer"."Epidemiológico".Nascimentos."Nascidos vivos por macrorregião de saúde"'

lb <- sqlQuery(channel, 
                  query, 
                  as.is = TRUE)

```

The SQL code used to create the query `'SELECT * FROM "Analytics Layer"."Epidemiológico".Nascimentos."Nascidos vivos por macrorregião de saúde"'` is presented below.

``` sql

WITH nascimentos_final 
AS (
    SELECT 
        SUBSTR(CODMUNRES, 1, 6) AS CODUFMUN,
        SUBSTR(DTNASC, 5, 4) AS ANO,
        SUBSTR(DTNASC, 3, 2) AS MES,
        COUNT(*) AS QUANTIDADE

    FROM Dados.sinasc.DN
    GROUP BY
        SUBSTR(CODMUNRES, 1, 6),
        SUBSTR(DTNASC, 5, 4),
        SUBSTR(DTNASC, 3, 2)
    ORDER BY
        ANO, MES),

nascimentos_preliminar 
AS (
    SELECT 
        SUBSTR(CODMUNRES, 1, 6) AS CODUFMUN,
        SUBSTR(DTNASC, 5, 4) AS ANO,
        SUBSTR(DTNASC, 3, 2) AS MES,
        COUNT(*) AS QUANTIDADE
    FROM 
        Dados.sinasc.preliminar
    WHERE 
      SUBSTR(DTNASC, 5, 4) > (SELECT MAX(SUBSTR(DTNASC, 5, 4)) 
      FROM Dados.sinasc.DN)
    GROUP BY
        SUBSTR(CODMUNRES, 1, 6),
        SUBSTR(DTNASC, 5, 4),
        SUBSTR(DTNASC, 3, 2))

SELECT
    m.regiao,
    m.uf_sigla,
    m.cod_macrorregiao,
    m.macrorregiao,
    dn.ano,
    dn.mes,
    sum(quantidade) as quantidade
FROM
  (SELECT * FROM nascimentos_final
   UNION ALL
   SELECT * FROM nascimentos_preliminar) dn
LEFT JOIN "Analytics Layer".Territorial."Municípios - Hierarquia Completa" m 
  ON CAST(dn.CODUFMUN AS INT) = m.cod_municipio
WHERE uf_sigla IS NOT NULL
GROUP BY 
    m.regiao,
    m.uf_sigla,
    m.cod_macrorregiao,
    m.macrorregiao,
    dn.ano,
    dn.mes
ORDER BY
    ANO, MES 
```

```{r include=FALSE}

# this chunk of code is used to transform our data formats before any analysis

lb <- lb |> 
        rename(br_region = regiao) |> 
        rename(state = uf_sigla) |> 
        rename(mc_code = cod_macrorregiao) |> 
        rename(year = ano) |> 
        rename(month = mes) |> 
        rename(qtt = quantidade)


lb$date <- as.Date(paste(lb$year, 
                         lb$month, "1", 
                         sep = "-"), 
                   format = "%Y-%m-%d")

lb$qtt <- as.numeric(lb$qtt)


lb <- lb |> mutate(year = year(date))

lb <- lb |> 
        filter(date < "2023-01-01")

```

# A) Case Study for one macro-region

First, we'll run the forecasting for just one health macro-region. Once we have consolidated this rationale, we can move on to automating the same steps for the other locations using functions. We randomly selected Centro-Sudeste health macro-region, which is located in the state of Goias, in the middle-west of Brazil.

The code below is used to plot the number of LB between 2000 and the end of 2022 (the last year with definitive data). We can see that in recent years this macro-region has seen a downward trend in the number of live births. In addition, it can be seen that seasonality is generally present each year, with a higher number of LB in the first few months, until may.

```{r}
#| warning: false

a <- lb |> 
  filter(mc_code == "5210") |> 
  ggplot(aes(x = date, y = qtt)) + 
  geom_line() + geom_smooth(se = FALSE) + 
  theme_minimal() + theme(legend.position = "bottom") + 
  xlab("Date") + ylab("LB Total") + 
  ggtitle("Evolution of LB - Centro-Sudeste(GO) macro-region")


b <- lb |> 
  filter(year == 2018 & mc_code == "5210") |> 
  ggplot(aes(x = date, y = qtt)) + 
  geom_line() + geom_smooth(se = FALSE) + 
  theme_minimal() + theme(axis.text.x = 
                            element_text(angle = 90, hjust = 1)) + 
  xlab("Date") + ylab("LB Total") + ggtitle("2018")

c <- lb |> 
  filter(year == 2019 & mc_code == "5210") |> 
  ggplot(aes(x = date, y = qtt)) + 
  geom_line() + geom_smooth(se = FALSE) + 
  theme_minimal() + theme(axis.text.x = 
                            element_text(angle = 90, hjust = 1)) + 
  xlab("Date") + ylab("LB Total") + ggtitle("2019")


d <- lb |> 
  filter(year == 2020 & mc_code == "5210") |> 
  ggplot(aes(x = date, y = qtt)) + 
  geom_line() + geom_smooth(se = FALSE) + 
  theme_minimal() + theme(axis.text.x = 
                            element_text(angle = 90, hjust = 1)) + 
  xlab("Date") + ylab("LB Total") + ggtitle("2020")

a / (b + c + d)

```

## 1) Time split

Let's do a time split to divide the data set into two subsets: training and test.

-   Training is a subset that supplies the algorithms;

-   Test is a subset used to assess how well the algorithm's predictive output adheres to real values.

80% of the data set will be used for training and 20% for testing.

```{r}

lb5210 <- lb |> filter(mc_code == "5210")
  
splits <- initial_time_split(lb5210, prop = 0.80)

splits
```

## 2) Fit models

We built an object containing various forecasting models, including Autoregressive Integrated Moving Average (ARIMA), ARIMA with XGBoost, Prophet, Exponential Smoothing (ETS) and Multivariate Adaptive Regression Splines (MARS).

```{r}
#| warning: false

# Model 1: auto_arima ----
model_fit_arima_no_boost <-
    arima_reg() |>
    set_engine(engine = "auto_arima") |>
    fit(qtt ~ date, data = training(splits))


# Model 2: arima_boost ----

model_fit_arima_boosted <- 
  arima_boost(
    min_n = 2,
    learn_rate = 0.015)  |> 
    set_engine(engine = "auto_arima_xgboost") |> 
    fit(qtt ~ date + 
              as.numeric(date) + 
              factor(month(date, 
                           label = TRUE),
                     ordered = F),
        data = training(splits))

# Model 3: ets ----
model_fit_ets <- 
  exp_smoothing() |>
    set_engine(engine = "ets") |>
    fit(qtt ~ date, data = training(splits))

# Model 4: prophet ----
model_fit_prophet <- 
  prophet_reg() |>
    set_engine(engine = "prophet") |>
    fit(qtt ~ date, data = training(splits))


# Model 5: lm ----
model_fit_lm <- 
  linear_reg() |>
    set_engine("lm") |>
    fit(qtt ~ as.numeric(date) + 
                factor(month(date, 
                             label = TRUE), 
                       ordered = FALSE),
        data = training(splits))

# Model 6: earth ----

model_spec_mars <- 
  mars(mode = "regression") |>
    set_engine("earth") 

recipe_spec <- 
  recipe(qtt ~ date, data = training(splits)) |>
    step_date(date, features = "month", ordinal = FALSE) |>
    step_mutate(date_num = as.numeric(date)) |>
    step_normalize(date_num) |>
    step_rm(date)
  
wflw_fit_mars <- 
  workflow() |>
    add_recipe(recipe_spec) |>
    add_model(model_spec_mars) |>
    fit(training(splits))


models_tbl <- 
  modeltime_table(
    model_fit_arima_no_boost,
    model_fit_arima_boosted,
    model_fit_ets,
    model_fit_prophet,
    model_fit_lm,
    wflw_fit_mars)

models_tbl

```

## 3) Model evaluation

For each of the models listed above, we have analyzed which models have the greatest predictive capacity. First, we just make a visual comparison.

```{r}

calibration_tbl <- 
    models_tbl |>
    modeltime_calibrate(new_data = testing(splits))

calibration_tbl |>
    modeltime_forecast(
        new_data  = testing(splits),
        actual_data = lb5210) |>
    plot_modeltime_forecast(.conf_interval_show = FALSE)

```

Next, we evaluated which models perform best using error metrics. Taking the Mean Absolut Percentage Error (MAPE) as a reference, the best model is the one with the lowest errors, i.e. ARIMA with XGBoost.

```{r}

calibration_tbl |>
    modeltime_accuracy() |>
    table_modeltime_accuracy()

min_mape <- calibration_tbl |> 
                modeltime_accuracy() |> 
                slice_min(mape, n = 1)


```

We retrain the model, now with all the data, including the test dataset. In addition, we apply a projection for the next 3 years using the best model. We saved the model and the projections to plot in a graph at the end.

```{r}

# Refit model using training and test dataset
refit_tbl <- 
    calibration_tbl |>
    modeltime_refit(data = lb5210)

# Projection of 3 years
refit_tbl |>
    modeltime_forecast(h = "3 years", 
                       actual_data = lb5210) |>
    plot_modeltime_forecast(
      .conf_interval_show = FALSE,
      .legend_max_width = 25)

# Creating a temporary object temp to save the best model's output

temp <- 
  refit_tbl |> 
  filter(.model_id == min_mape$.model_id) |> 
  modeltime_forecast(h = "3 years", 
                       actual_data = lb5210) |>
    plot_modeltime_forecast(
      .conf_interval_show = FALSE,
      .legend_max_width = 25)

date_i <- temp[["x"]][["data"]][[1]][["x"]]
qtt_i <- temp[["x"]][["data"]][[1]][["y"]]   
    
date <- temp[["x"]][["data"]][[2]][["x"]]
qtt <- temp[["x"]][["data"]][[2]][["y"]]

state_region <- lb |> 
                  filter(mc_code == "5210")

state <- unique(state_region$state)
region_name <- unique(state_region$macrorregiao)
mc_region <- unique(state_region$mc_code)

df_real <- tibble(mc_region = "5210",
                  region_name = region_name,
                  state = state,
                  mape = "real data",
                  id = "real data",
                  desc = "real data",
                  date = date_i,
                  qtt = round(qtt_i))

df_projection <- tibble(mc_region = "5210",
                        region_name = region_name,
                        state = state,
                        mape = min_mape$mape,
                        id = min_mape$.model_id,
                        desc = min_mape$.model_desc,
                        date,
                        qtt)

# creating an object with real data and projected data

df <- rbind(df_real,
            df_projection)

# saving the object into a directory

filename <- 
  sprintf("output %s - %s - %s.csv", 
          mc_region, 
          state, 
          region_name)

filepath <- 
  file.path("~/GitHub/Tese/Tese/01_script/forecast_files",
            filename)

write.csv(df, filepath, row.names = FALSE)


# Plotting forecast

graph <- 
  df |> 
  ggplot(aes(x = date, 
             y = qtt, 
             col = desc)) +
  geom_line() + 
  theme_minimal() + 
  ggtitle(sprintf("Forecasting - %s - %s - %s", 
                  mc_region, state, region_name)) +
  theme(plot.background = 
          element_rect(fill = "white", 
                       color = NA),
        panel.background = 
          element_rect(fill = "white", 
                       color = NA),
        legend.position = "bottom")

graph

# Saving graph in .png format 

  
ggsave(filename = sprintf("Graph %s - %s - %s.png", 
                          mc_region, 
                          state,
                          region_name), 
     plot = graph, 
     path = "~/GitHub/Tese/Tese/01_script/graph",
     width = 10, height = 5)


```

# B) Function to automate process

We created a function to apply and maintain the best trained model based on the experience expressed in section A.

```{r}
#| warning: FALSE

best_model <- 
  
  function(mc_region){
    
    # split
    lb_region <- lb |> 
      filter(mc_code == mc_region)
    
    splits <- initial_time_split(lb_region, 
                                 prop = 0.80)
    
    # models 
    
      # Model 1: auto_arima ----
      model_fit_arima_no_boost <-
          arima_reg() |>
          set_engine(engine = "auto_arima") |>
          fit(qtt ~ date, data = training(splits))
      
  
      # Model 2: arima_boost ----
      model_fit_arima_boosted <- 
        arima_boost(
          min_n = 2,
          learn_rate = 0.015)  |> 
          set_engine(engine = "auto_arima_xgboost") |> 
          fit(qtt ~ date + 
                    as.numeric(date) + 
                    factor(month(date, 
                                 label = TRUE),
                           ordered = F),
              data = training(splits))
      
      # Model 3: ets ----
      model_fit_ets <- exp_smoothing() |>
          set_engine(engine = "ets") |>
          fit(qtt ~ date, data = training(splits))
      
      # Model 4: prophet ----
      model_fit_prophet <- prophet_reg() |>
          set_engine(engine = "prophet") |>
          fit(qtt ~ date, data = training(splits))
      
    
    # table with models
      
    models_tbl <- 
      modeltime_table(
        model_fit_arima_no_boost,
        model_fit_arima_boosted,
        model_fit_ets,
        model_fit_prophet)
    
    calibration_tbl <- 
      models_tbl |>
      modeltime_calibrate(new_data = testing(splits))
    
    # evaluating the best model 
    
    min_mape <- calibration_tbl |> 
                modeltime_accuracy() |> 
                slice_min(mape, n = 1)
    
    refit_tbl <- calibration_tbl |>
                    modeltime_refit(data = lb_region) |>
                    filter(.model_id == min_mape$.model_id)
    
temp <- 
  refit_tbl |> 
  filter(.model_id == min_mape$.model_id) |> 
  modeltime_forecast(h = "3 years", 
                       actual_data = lb_region) |>
    plot_modeltime_forecast(
      .conf_interval_show = FALSE,
      .legend_max_width = 25)

date_i <- temp[["x"]][["data"]][[1]][["x"]]
qtt_i <- temp[["x"]][["data"]][[1]][["y"]]   
    
date <- temp[["x"]][["data"]][[2]][["x"]]
qtt <- temp[["x"]][["data"]][[2]][["y"]]

state_region <- lb_region |> 
                  filter(mc_code == mc_region)

state <- unique(state_region$state)
region_name <- unique(state_region$macrorregiao)
mc_region <- unique(state_region$mc_code)

df_real <- tibble(mc_region = mc_region,
                  region_name = region_name,
                  state = state,
                  mape = "real data",
                  id = "real data",
                  desc = "real data",
                  date = date_i,
                  qtt = round(qtt_i))

df_projection <- tibble(mc_region = mc_region,
                        region_name = region_name,
                        state = state,
                        mape = min_mape$mape,
                        id = min_mape$.model_id,
                        desc = min_mape$.model_desc,
                        date,
                        qtt)

# creating an object with real data and projected data

df <- rbind(df_real,
            df_projection)
  
# saving the output file of projections

filename <- 
  sprintf("%s - %s.csv", 
          mc_region, 
          state, 
          region_name)

filepath <- 
  file.path("~/GitHub/Tese/Tese/01_script/forecast_files",
            filename)

write.csv(df, filepath, row.names = FALSE)

return(df)

}

```

The following function plots the graph and saves it into a proper directory.

```{r}

grafico <- function(df){
  
  df <- df
  
  region_code = df$mc_region
  region_name = df$region_name
  region_state = df$state
  
  graph <- 
    df |> 
    ggplot(aes(x = date, 
               y = qtt, 
               col = desc)) +
    geom_line() + 
    theme_minimal() + 
    ggtitle(sprintf("Forecasting %s - %s - %s.png", 
                            region_code, 
                            region_state,
                            region_name)) +
    theme(plot.background = 
            element_rect(fill = "white", 
                         color = NA),
          panel.background = element_rect(fill = "white", 
                                          color = NA),
          legend.position = "bottom")

# saving the graphs in a proper directory
    
  ggsave(filename = sprintf("Graph %s - %s - %s.png", 
                            region_code, 
                            region_state,
                            region_name), 
       plot = graph, 
       path = "~/GitHub/Tese/Tese/01_script/graph",
       width = 10, height = 5)
  
  graph

}

```

# C) Forecasting all regions

Using a looping structure, we automate the process using both functions from section B.

```{r eval=FALSE}
#| warning: FALSE

region_vector <- lb |> 
                  distinct(mc_code)  
    

region_vector <- region_vector$mc_code

df_total <- data.frame()

for(i in region_vector){
  
  df <- best_model(i)
  
  df_total <- rbind(df_total, df)
  
  grafico(df)
  
}

```

```{r eval=FALSE, include=FALSE}


df_goias <- df_total
df_mt <- df_total
df_ms <- df_total
df_to <- df_total
df_ap <- df_total
df_ac <- df_total
df_rr <- df_total
df_ro <- df_total
df_pa <- df_total
df_pr <- df_total
df_sc <- df_total
df_rs <- df_total
df_df <- df_total
df_am <- df_total
df_ba <- df_total
df_rn_se_pe <- df_total
df_ma <- df_total
df_pi <- df_total
df_al <- df_total
df_pb <- df_total
df_es <- df_total
df_rj <- df_total
df_sp <- df_total
df_mg <- df_total
df_ce_pb <- df_total



df_geral <- 
  rbind(df_goias, df_mt, df_ms, df_df,
        df_to, df_pa, df_am, df_ap, df_rr, df_ro, df_ac, df_ap,
        df_mg, df_sp, df_rj, df_es, 
        df_rs, df_sc, df_pr, 
        df_rn_se_pe, df_ba, df_pi, df_ce_pb, df_ma, df_al)

#write.csv(df_geral, "df_geral_previsoes.csv")

```

```{r include=FALSE}

df_geral <- read_csv("~/GitHub/Tese/Tese/01_script/forecast_files/df_geral_previsoes.csv")

df_total <- df_geral

```

# D) Final considerations

Using the modeltime framework we automated the live birth forecast for 113 macro-regions in Brazil. We have acchieved low level errors. The following boxplot shows that the median percentage error is 5.62. Additionally, the third quantile corresponded to a percentage error of 7.24. Only six macro-regions had percentage errors between 10 and 14.

```{r}
mape <- 
  df_total |> 
  filter(mape != "real data")

mape$mape <- as.numeric(mape$mape)

mape |> 
  ggplot(aes(x = mape)) + geom_boxplot() + 
  coord_flip() + theme_minimal()

  
  
```
