---
title: "Forecasting live births"
format: html
editor: visual
---

This script aims to describe the process of applying the method for forecasting the number of live births (LB) in each health macro-region in Brazil. To do this, we will use data from the Live Births System (SINASC, *Sistema de Nascidos Vivos*) provided by the SUS Information Technology Department (DATASUS) for each health macro-region.

# Preparing R environment

Loading libraries needed to forecast LB.

```{r}
#| warning: FALSE

library(tidyverse)
library(modeltime)
library(RODBC)
library(xgboost)
library(tidymodels)
library(timetk)
library(patchwork)

```

# Data extraction

We loaded the DATASUS data into our own infrastructure in order to access it using SQL, given the volume of data and the need to increase storage and processing efficiency.

The script below performs the data access.

```{r}
#| warning: false

dremio_host <- Sys.getenv("endereco")
dremio_port <- Sys.getenv("port")
dremio_uid <- Sys.getenv("uid")
dremio_pwd <- Sys.getenv("datalake")

channel <- 
        odbcDriverConnect(
                sprintf("DRIVER=Dremio Connector;
                        HOST=%s;PORT=%s;UID=%s;PWD=%s;
                        AUTHENTICATIONTYPE=Basic Authentication;
                        CONNECTIONTYPE=Direct", 
                        dremio_host, dremio_port, 
                        dremio_uid, dremio_pwd))

query <- 'SELECT * 
            FROM "Analytics Layer"."Epidemiológico".Nascimentos."Nascidos vivos por macrorregião de saúde"'

lb <- sqlQuery(channel, 
                  query, 
                  as.is = TRUE)

```

The SQL code used to create the query `'SELECT * FROM "Analytics Layer"."Epidemiológico".Nascimentos."Nascidos vivos por macrorregião de saúde"'` is presented below.

``` sql

WITH nascimentos_final 
AS (
    SELECT 
        SUBSTR(CODMUNRES, 1, 6) AS CODUFMUN,
        SUBSTR(DTNASC, 5, 4) AS ANO,
        SUBSTR(DTNASC, 3, 2) AS MES,
        COUNT(*) AS QUANTIDADE

    FROM Dados.sinasc.DN
    GROUP BY
        SUBSTR(CODMUNRES, 1, 6),
        SUBSTR(DTNASC, 5, 4),
        SUBSTR(DTNASC, 3, 2)
    ORDER BY
        ANO, MES),

nascimentos_preliminar 
AS (
    SELECT 
        SUBSTR(CODMUNRES, 1, 6) AS CODUFMUN,
        SUBSTR(DTNASC, 5, 4) AS ANO,
        SUBSTR(DTNASC, 3, 2) AS MES,
        COUNT(*) AS QUANTIDADE
    FROM 
        Dados.sinasc.preliminar
    WHERE 
      SUBSTR(DTNASC, 5, 4) > (SELECT MAX(SUBSTR(DTNASC, 5, 4)) 
      FROM Dados.sinasc.DN)
    GROUP BY
        SUBSTR(CODMUNRES, 1, 6),
        SUBSTR(DTNASC, 5, 4),
        SUBSTR(DTNASC, 3, 2))

SELECT
    m.regiao,
    m.uf_sigla,
    m.cod_macrorregiao,
    m.macrorregiao,
    dn.ano,
    dn.mes,
    sum(quantidade) as quantidade
FROM
  (SELECT * FROM nascimentos_final
   UNION ALL
   SELECT * FROM nascimentos_preliminar) dn
LEFT JOIN "Analytics Layer".Territorial."Municípios - Hierarquia Completa" m 
  ON CAST(dn.CODUFMUN AS INT) = m.cod_municipio
WHERE uf_sigla IS NOT NULL
GROUP BY 
    m.regiao,
    m.uf_sigla,
    m.cod_macrorregiao,
    m.macrorregiao,
    dn.ano,
    dn.mes
ORDER BY
    ANO, MES 
```

```{r include=FALSE}

# this chunk of code is used to transform our data formats before any analysis

lb <- lb |> 
        rename(br_region = regiao) |> 
        rename(state = uf_sigla) |> 
        rename(mc_code = cod_macrorregiao) |> 
        rename(year = ano) |> 
        rename(month = mes) |> 
        rename(qtt = quantidade)


lb$date <- as.Date(paste(lb$year, 
                         lb$month, "1", 
                         sep = "-"), 
                   format = "%Y-%m-%d")

lb$qtt <- as.numeric(lb$qtt)


lb <- lb |> mutate(year = year(date))

lb <- lb |> 
        filter(date < "2023-01-01")

```

# A) Case Study for one macro-region

First, we'll run the forecasting for just one health macro-region. Once we have consolidated this rationale, we can move on to automating the same steps for the other locations using functions. We randomly selected Centro-Sudeste health macro-region, which is located in the state of Goias, in the middle-west of Brazil.

The code below is used to plot the number of LB between 2000 and the end of 2022 (the last year with definitive data). We can see that in recent years this macro-region has seen a downward trend in the number of live births. In addition, it can be seen that seasonality is generally present each year, with a higher number of LB in the first few months of the year.

```{r}
#| warning: false

a <- lb |> 
  filter(mc_code == "5210") |> 
  ggplot(aes(x = date, y = qtt)) + 
  geom_line() + geom_smooth(se = FALSE) + 
  theme_minimal() + theme(legend.position = "bottom") + 
  xlab("Date") + ylab("LB Total") + 
  ggtitle("Evolution of LB - Centro-Sudeste(GO) macro-region")


b <- lb |> 
  filter(year == 2018 & mc_code == "5210") |> 
  ggplot(aes(x = date, y = qtt)) + 
  geom_line() + geom_smooth(se = FALSE) + 
  theme_minimal() + theme(axis.text.x = 
                            element_text(angle = 90, hjust = 1)) + 
  xlab("Date") + ylab("LB Total") + ggtitle("2018")

c <- lb |> 
  filter(year == 2019 & mc_code == "5210") |> 
  ggplot(aes(x = date, y = qtt)) + 
  geom_line() + geom_smooth(se = FALSE) + 
  theme_minimal() + theme(axis.text.x = 
                            element_text(angle = 90, hjust = 1)) + 
  xlab("Date") + ylab("LB Total") + ggtitle("2019")


d <- lb |> 
  filter(year == 2020 & mc_code == "5210") |> 
  ggplot(aes(x = date, y = qtt)) + 
  geom_line() + geom_smooth(se = FALSE) + 
  theme_minimal() + theme(axis.text.x = 
                            element_text(angle = 90, hjust = 1)) + 
  xlab("Date") + ylab("LB Total") + ggtitle("2020")

a / (b + c + d)

```

## 1) Time split

Let's do a time split to divide the data set into two subsets: training and test.

-   Training is a subset that supplies the algorithms;

-   Test is a subset used to assess how well the algorithm's predictive output adheres to real values.

80% of the data set will be used for training and 20% for testing.

```{r}

lb5210 <- lb |> filter(mc_code == "5210")
  
splits <- initial_time_split(lb5210, prop = 0.80)

splits
```

## 2) Fit models
